An Investigation of Confusing Code Patterns in JavaScript

### 2023-02-21 abstract.tex
endereçando os comentários de R1 e R3 após o segundo round de revisões

\begin{abstract}
Evolving software is particularly challenging when the code has been poorly written or
uses confusing idioms and language constructs, which might increase maintenance efforts
and impose a significant cognitive load on developers. Previous research has
investigated possible sources of confusion in programs, including the impact of small code
patterns (hereafter atoms of confusion) that contribute to misunderstanding the source code.
Although researchers have explored atoms of confusion in code written in \clang, \cpplang,
and Java, different languages have different features, developer communities, and development cultures. This
justifies the exploration of other languages to verify whether they also exhibit confusion-inducing patterns. 
In this paper we investigate the impact of atoms of confusion on understanding JavaScript code---a
dynamically typed language whose popularity is growing in the most diverse application domains. In this paper
we report the results of a mixed-methods research comprising a mining software repositories (MSR) study, two experiments, and a set of interviews with practitioners.  
Our MSR effort shows that atom candidastes are frequent and used intensively in 72 popular open-source JavaScript projects:
four atom candidates appear in 90\% of them and two of them occur more than once for every 100 lines of code. This helps motivate the other three studies. 
The results of both experiments suggest that two code patterns that have been previously observed to confuse \clang
programmers also confuse JavaScript programmers: the comma operator and assignments
being used as values. In addition, some code patterns, such as omitted curly braces and change of literal encoding, have caused confusion in participants in one of the experiments. We discover that some JavaScript-specific elements, such as automatic semicolon insertion and object destructuring, also have the potential to cause confusion. 
%In addition, at least one experiment has confirmed that some atoms detected for other languages, such as  and type conversion, also 
%In addition, pre- and post-increment operators, when used as values, omitted curly braces, changes of literal encoding, and type conversions have been confirmed as confuse in at least one
%of the two experiments. 
For all these cases effect sizes were either medium or high. The interviews we conducted
%reinforce some of these findings while pointing to
indicate other constructs and idioms that merit investigation in the future. 
%Our findings provide significant evidence that some code patterns should be avoided if code is expected to be understood by a wide audience. 
\end{abstract}

>>>>

\begin{abstract}
Evolving software is particularly challenging when the code has been poorly written or
uses confusing idioms and language constructs, which might increase maintenance efforts
and impose a significant cognitive load on developers. Previous research has
investigated possible sources of confusion in programs, including the impact of small code
patterns (hereafter atoms of confusion) that contribute to misunderstanding the source code.
Although researchers have explored atoms of confusion in code written in \clang, \cpplang,
and Java, different languages have different features, developer communities, and development cultures. This
justifies the exploration of other languages to verify whether they also exhibit confusion-inducing patterns. 
In this paper we investigate the impact of atoms of confusion on understanding JavaScript code---a
dynamically typed language whose popularity is growing in the most diverse application domains. We present
the results of a mixed-methods research comprising a mining software repositories (MSR) study, two experiments, and a set of interviews with practitioners.  
Our MSR effort shows that atom candidastes are frequent and used intensively in 72 popular open-source JavaScript projects:
four atom candidates appear in 90\% of them and two of them occur more than once for every 100 lines of code. This helps motivate the other three studies. 
The results of both experiments suggest that two code patterns that have been previously observed to confuse \clang
programmers also confuse JavaScript programmers: the comma operator and assignments
being used as values. In addition, some code patterns, such as omitted curly braces and change of literal encoding, have caused confusion in participants in one of the experiments. We discover that some JavaScript-specific elements, such as automatic semicolon insertion and object destructuring, also have the potential to cause confusion. 
%In addition, at least one experiment has confirmed that some atoms detected for other languages, such as  and type conversion, also 
%In addition, pre- and post-increment operators, when used as values, omitted curly braces, changes of literal encoding, and type conversions have been confirmed as confuse in at least one
%of the two experiments. 
For all these cases effect sizes were either medium or high. The interviews we conducted
%reinforce some of these findings while pointing to
indicate other constructs and idioms that merit investigation in the future. 
%Our findings provide significant evidence that some code patterns should be avoided if code is expected to be understood by a wide audience. 
\end{abstract}

### 2023-02-21 discussion.tex
endereçando os comentários de R1 e R3 após o segundo round de revisões

In the repeated measures study, we identified one JavaScript-specific atom, Object Destructuring. In the Latin square study, Automatic Semicolon Insertion was identified as another JavaScript-specific atom. For both cases, p-values were much lower than the threshold and effect sizes were high. However, in the repeated measures study obfuscated versions of Automatic Semicolon Insertion were not significantly different from their clean counterparts. Our interviewees favored the use of semicolon over Automatic Semicolon Insertion. Finally, as discussed in Section~\ref{sec:repeated:correct}, Array Destructuring exhibited a large effect size, in spite of the non-statistically significant difference (after correction). As discussed elsewhere~\cite{Ellis:2010:EGE}, this is a scenario that hints at practical relevance, especially considering the very low p-value ($< 0.003$). Further studies investigating these code patterns, potentially using different response variables~\cite{Oliveira:2020:ECR,Feitelson:2021:CPC}, are left for future work.

Some code patterns that have been identified as atoms in previous work were not confirmed in our experiments, at least from a correctness perspective. Comparing the results of the Latin square study to previous research, two atoms of confusion that Gopstein et al.~\cite{DBLP:conf/sigsoft/GopsteinIYDZYC17} confirmed for \clang and \cpplang do not lead to a statistically significant impact on correctness in our study, Logic as Control Flow, and Conditional Operator. Conditional Operator was the only atom candidate in our interviews whose obfuscated version was considered less confusing than the corresponding clean version.

Logic as Control Flow is a more interesting case. Considering the correctness perspective, none of our experiments has identified Logic as Control Flow as an atom of confusion. However, in the Latin square study there was a significant difference in the time required to make a correct prediction in favor of the clean version of this atom candidate, with a medium effect size. Furthermore, both Langhout and Aniche~\cite{Langhout:2021:ACJ} and Gopstein and colleagues~\cite{DBLP:conf/sigsoft/GopsteinIYDZYC17} have identified it as an atom. Two other atoms have exhibited a statistically significant difference in time, in favor of their obfuscated versions (with a small effect size). One of them, Pre-Increment, has been confirmed as an atom of confusion in multiple experiments, i.e., participants are less confused by the clean versions. These conflicting results reinforce the importance of conducting experiments that consider multiple response variables~\cite{Oliveira:2020:ECR}. In addition, it is clear that factors such as the experience of the participants and the selection of code snippets can have considerable impact on the results of studies about code readability~\cite{Feitelson:2021:CPC}.

@@

Altogether, we answer our first research question as follows.

\begin{mh}
  {\bf Answer to RQ1:} Our experiments using repeated measures and the Latin square designs give evidence that some of the atom candidates in \clang and \cpplang programs that also exist in JavaScript correspond to a source of misunderstanding in
  JavaScript code. 
\end{mh}

@@

\begin{mh}
  {\bf Answer to RQ2:} The qualitative analysis of the
  interviews supports the results of our experiments---
  JavaScript developers most often agree that atoms of confusion compromise
  source code understanding. 
\end{mh}

@@

The results of the third study (mining open source
JavaScript repositories) provides evidence that,
although atoms of confusion compromise program
comprehension, they frequently appear in open
source JavaScript projects. In particular,
seven, out of 10 atoms considered
  in our study, appear in more than 50\% of
the 72 projects we analyzed. Furthermore, at least two of them are used intensively, more than once for every 200 lines of code. In summary, the third study
allows us to answer the third research
question.

\begin{mh}
  {\bf Answer to RQ3:} The MSR study reveals that
  several atom candidates explored in our research
  appear frequently in practice, and cleaning up the use of 
  Post-Increment/Decrement and the Automatic Semicolon Insertion
  might improve the readabiliy of JavaScript code substantially. 
\end{mh}

>>>>

In the repeated measures study, we identified one JavaScript-specific atom, \textbf{Object Destructuring}. In the Latin square study, \textbf{Automatic Semicolon Insertion} was identified as another JavaScript-specific atom. For both cases, p-values were much lower than the threshold and effect sizes were high. However, in the repeated measures study obfuscated versions of Automatic Semicolon Insertion were not significantly different from their clean counterparts. Our interviewees favored the use of semicolon over Automatic Semicolon Insertion. Finally, as discussed in Section~\ref{sec:repeated:correct}, \textbf{Array Destructuring} exhibited a large effect size, in spite of the non-statistically significant difference (after correction). As discussed elsewhere~\cite{Ellis:2010:EGE}, this is a scenario that hints at practical relevance, especially considering the very low p-value ($< 0.003$). Further studies investigating these code patterns, potentially using different response variables~\cite{Oliveira:2020:ECR,Feitelson:2021:CPC}, are left for future work.

Some code patterns that have been identified as atoms in previous work were not confirmed in our experiments, at least from a correctness perspective. Comparing the results of the Latin square study to previous research, two atoms of confusion that Gopstein et al.~\cite{DBLP:conf/sigsoft/GopsteinIYDZYC17} confirmed for \clang and \cpplang do not lead to a statistically significant impact on correctness in our study, \textbf{Logic as Control Flow}, and \textbf{Conditional Operator}. \textbf{Conditional Operator} was the only atom candidate in our interviews whose obfuscated version was considered less confusing than the corresponding clean version.

\textbf{Logic as Control Flow} is a more interesting case. Considering the correctness perspective, none of our experiments has identified Logic as Control Flow as an atom of confusion. However, in the Latin square study there was a significant difference in the time required to make a correct prediction in favor of the clean version of this atom candidate, with a medium effect size. Furthermore, both Langhout and Aniche~\cite{Langhout:2021:ACJ} and Gopstein and colleagues~\cite{DBLP:conf/sigsoft/GopsteinIYDZYC17} have identified it as an atom. Two other atoms have exhibited a statistically significant difference in time, in favor of their obfuscated versions (with a small effect size). One of them, \textbf{Pre-Increment}, has been confirmed as an atom of confusion in multiple experiments, i.e., participants are less confused by the clean versions. These conflicting results reinforce the importance of conducting experiments that consider multiple response variables~\cite{Oliveira:2020:ECR}. In addition, it is clear that factors such as the experience of the participants and the selection of code snippets can have considerable impact on the results of studies about code readability~\cite{Feitelson:2021:CPC}. \revised{Table~\ref{tab:summary-of-findings} summarizes the correctness assessment results in our two experiments (columns Exp01 and Exp02) and the experiments of
Gopstein et al.~\cite{DBLP:conf/sigsoft/GopsteinIYDZYC17} and Langhout and Aniche~\cite{Langhout:2021:ACJ}.}

\begin{table}[htb]
  \caption{Summary of the findings accross the different studies. Here we show 
    only the atom candidates that have been explored in at least one
    of our experiments. A {\bf True} value means that the experiment
    confirms the code pattern as an atom candidate ({\bf False} otherwise).}
  \label{tab:summary-of-findings}
\begin{scriptsize}
\begin{tabular}{lcccc}\toprule 
Atom Candidate	                 & Gopstein et al.~\cite{DBLP:conf/sigsoft/GopsteinIYDZYC17} & Langhout and Aniche~\cite{Langhout:2021:ACJ}  & Exp01     &  Exp02   \\ \midrule

Change of Literal Encoding       &   TRUE   &   TRUE   &   TRUE    &   -      \\
Assignment as Value              &   TRUE   &   -      &   TRUE    &   TRUE   \\
Logic as Control Flow            &   TRUE   &   TRUE   &   FALSE   &   FALSE  \\
Post-Increment                   &   TRUE   &   TRUE   &   FALSE   &   TRUE   \\
Type Conversion                  &   TRUE   &   TRUE   &   TRUE    &   -      \\
Conditional Operator             &   TRUE   &   FALSE  &   FALSE   &   FALSE  \\
Infix Operator Precedence        &   TRUE   &   FALSE  &   FALSE   &   -      \\
Comma Operator                   &   TRUE   &   -      &   TRUE    &   TRUE   \\
Pre-Increment                    &   TRUE   &   TRUE   &   TRUE    &   FALSE  \\
Implicit Predicate               &   TRUE   &   -      &   FALSE   &   TRUE   \\
Repurposed Variable              &   TRUE   &   FALSE  &   FALSE   &   -      \\ 
Omitted Curly Braces             &   TRUE   &   TRUE   &   TRUE    &   FALSE  \\
Dead Unreachable Repeated        &   FALSE  &   FALSE  &   FALSE   &   -      \\
Arithmetic as Logic              &   FALSE  &   FALSE  &   TRUE    &   FALSE  \\
Object Destructuring             &   -      &   -      &   TRUE    &   -      \\
Lack of Indentation no Braces    &   -      &   TRUE   &   TRUE    &   -      \\
Array Destructuring              &   -      &   -      &   FALSE   &   -      \\
Array Spread                     &   -      &   -      &   FALSE   &   -      \\
Object Spread                    &   -      &   -      &   FALSE   &   -      \\
Property Access                  &   -      &   -      &   FALSE   &   -      \\
Arrow Function                   &   -      &   -      &   FALSE   &   -      \\
Automatic Semicolon Insertion    &   -      &   -      &   FALSE   &   TRUE   \\
Lack of Indentation with Braces  &   -      &   -      &   FALSE   &   -      \\ \bottomrule

\end{tabular}
\end{scriptsize}
\end{table}

\revised{It is not feasible to combine the results of the statistical assessments coming from our two experiments---because their design differs significantly.
  Nonetheless, we can reason about the relationships between correct/wrong answers in both experiments and combine the results of the exploratory data analysis
  we presented in Sections~\ref{sec:repeated:correct} and~\ref{sec:latin:correct}.
  Considering the ten atoms explored in our two experiments, for six atom candidates the number of correct answers for the clean
  version of the code snippets is at least 25\% higher than the number of correct answers for the obfuscated version of the code (see Table~\ref{tab:merge-exploratory-study}).}

\begin{table}
  \caption{Results of the exploratory data analysis for the experiments (Repeated Measurea and Latin Square)}
  \label{tab:merge-exploratory-study}
  \begin{scriptsize}
  \begin{tabular}{lccccc} \toprule
                                 & \multicolumn{2}{c}{Repeated Measures} & \multicolumn{2}{c}{Latin Square} & \\  
  Atom Candidate                 &   Obfuscated   &   Clean   &   Obfuscated   &   Clean   &  $\delta$ (\%)    \\ \midrule
Comma Operator                   &      8         &   29      &       28       &     65    &   161.11  \\
Automatic Semicolong Insertion   &     15         &   21      &       32       &     68    &    89.36  \\
Assignment as Value              &     18         &   43      &       56       &     68    &    50.00  \\
Post-Increment                   &      8         &   17      &       48       &     64    &    44.64  \\
Omitted Curly Braces             &     35         &   51      &       47       &     58    &    32.92  \\
Pre-Increment                    &     13         &   28      &       50       &     53    &    28.57  \\
Implicit Predicate               &     41         &   44      &       58       &     68    &    13.13  \\
Conditional Operator             &     60         &   62      &       60       &     66    &     6.66  \\
Logic as Control Flow            &     33         &   28      &       41       &     48    &     2.70  \\ 
Arithmetic as Logic              &     52         &   47      &       64       &     63    &    -5.17  \\ \bottomrule
  \end{tabular}
\end{scriptsize}  
\end{table}

\revised{Besides that, the results of the two experiments reveal findings that diverge. That is, for six atom candidates (out of ten explored in both experiments),
  the results of the statistical tests differ. Three main reasons might explain the differences: (a) the code snippets used in the experiments are different,
  (b) the participants of the experiments are different, and (c) the design of the experiments and thus the analysis procedures are different.
  As such, additional research would be necessary to understand the impact of these items on the results of empirical assessments of atoms of confusion in general.}

\revised{Our findings here suggest the need for replication studies in this field. Here we left an additional question:
  What would be the results of the original studies on atoms of confusion if an independent group reproduces the original research? Although the divergences in the two experiments might threaten our study's conclusions, we complement the experiments with a qualitative study based on interviews. We return to this issue of reporting the results
of different experiments in Section~\ref{{sec:whytwo}}.}

@@

The results of the first study (mining open source
JavaScript repositories) provides evidence that,
although atoms of confusion compromise program
comprehension, they frequently appear in open
source JavaScript projects. In particular,
seven, out of 10 atoms considered
  in our study, appear in more than 50\% of
the 72 projects we analyzed. Furthermore, at least two of them are used intensively, more than once for every 200 lines of code. In summary, the first study
allows us to answer the third research
question.

\begin{mh}
  {\bf Answer to RQ1:} The MSR study reveals that
  several atom candidates explored in our research
  appear frequently in practice, and cleaning up the use of 
  Post-Increment/Decrement and the Automatic Semicolon Insertion
  might improve the readabiliy of JavaScript code substantially. 
\end{mh}

Altogether, the results of our experiments allow us
to answer the second research question of this
study as follows.

\begin{mh}
  {\bf Answer to RQ2:} Our experiments using repeated measures and the Latin square designs give evidence that some of
  the atom candidates in \clang and \cpplang programs that also exist in JavaScript correspond to a source of misunderstanding in
  JavaScript code. Two JavaScript-specific code patterns hava been identified as atoms of confusion by at least
  one of our experiments: Object Descructuring and Automatic Semicolon Insertion. 
\end{mh}

@@

\begin{mh}
  {\bf Answer to RQ3:} The qualitative analysis of the
  interviews supports the results of our experiments---
  JavaScript developers most often agree that atoms of confusion compromise
  source code understanding. 
\end{mh}

@@

### 2023-02-21 methodology.tex
endereçando os comentários de R1 e R3 após o segundo round de revisões

\revised{Answering RQ1 enables us to better motivate the other two questions by showing that studying these code patterns has practical relevance.}
Futhermore, by addressing research questions RQ2 and RQ3, we can either generalize or refute the findings about atoms of confusion already discussed in the literature. 
In addition, answering the third question allows us to enrich existing catalogs about atoms of confusion and discuss how often they occur in practice. Answering these research questions also lays the foundations for the implementation of tools that can automatically transform code into cleaner versions.
%---though we postpone these results to future research work.

We conduct a mixed-methods study to answer these questions. 
It includes two independently designed and conducted experiments: a repeated measures/within subjects study (Section~\ref{sec:s01}) and one using a Latin square, between subjects, counterbalancing experimental design to control for subject experience (Section~\ref{sec:s02}). By using two different experiments, conceived by different researchers, with different designs, and conducted with different participants, we hope to rigorously verify the impact of atoms of confusion in code comprehension activities. \revised{Conducting multiple studies, with different designs and samples, is a typical approach in other areas of science to obtain more solid evidence and also to contest evidence obtained by single study, for example, Weggemans et al.~\cite{Weggemans:2001:DCE}. We discuss this matter further in Section~\ref{sec:whytwo}}.

\revised{We perform a repository mining study (Section~\ref{sec:s04}) that investigates the prevalence of some of these atom candidates in large scale, professionally-developed JavaScript software. This helps us motivate the remaining three studies.}
These two experiments explore the impact of atom candidates on understanding JavaScript code (goal of research question RQ1). They investigate whether or not programs that contain atom candidates tend to produce more misunderstanding for programmers trying to predict their outcome.% We leverage two independently-developed and conducted experiments with different designs to more  
We also conduct a set of interviews with developers (Section~\ref{sec:s03}) to contrast the quantitative results of the experiments with their preferences and opinions (thus addressing RQ2). %Finally, we perform a repository mining study (Section~\ref{sec:s04}) that investigates the prevalence of some of these atom candidates in large scale, professionally-developed JavaScript software.

Table~\ref{tab:atomsBoth} presents the atom candidates that we explore in all stduies, while Table~\ref{tab:atomsRepeated} presents the additional 
ones that only the repeated measures study analyzes.

>>>>

\revised{Answering RQ1 enables us to better motivate the other two questions by showing that studying these code patterns has practical relevance and discussing how often they occur in practice.}
Futhermore, by addressing research questions RQ2 and RQ3, we can either generalize or refute the findings about atoms of confusion already discussed in the literature. 
In addition, answering the third question allows us to enrich existing catalogs about atom candidates. Answering these research questions might also lay the foundations for the implementation of tools that can automatically transform code into cleaner versions.
%---though we postpone these results to future research work.

We conduct a mixed-methods study to answer these questions.
\revised{We perform a repository mining study (Section~\ref{sec:s04}) that investigates the prevalence of some of these atom candidates in large scale, professionally-developed JavaScript software. This helps us motivate the remaining three studies.}
Our research also includes two independently designed and conducted experiments: a repeated measures/within subjects study (Section~\ref{sec:s01}) and one using a Latin square, between subjects, counterbalancing experimental design to control for subject experience (Section~\ref{sec:s02}). By using two different experiments, conceived by different researchers, with different designs, and conducted with different participants, we hope to rigorously verify the impact of atoms of confusion in code comprehension activities. \revised{Conducting multiple studies, with different designs and samples, is a typical approach in other areas of science to obtain more solid evidence and also to contest evidence obtained by single study, for example, Weggemans et al.~\cite{Weggemans:2001:DCE}. We discuss this matter further in Section~\ref{sec:whytwo}}.

These two experiments explore the impact of atom candidates on understanding JavaScript code (goal of research question RQ2). They investigate whether or not programs that contain atom candidates tend to produce more misunderstanding for programmers trying to predict their outcome.% We leverage two independently-developed and conducted experiments with different designs to more  
We also conduct a set of interviews with developers (Section~\ref{sec:s03}) to contrast the quantitative results of the experiments with their preferences and opinions (thus addressing RQ3). %Finally, we perform a repository mining study (Section~\ref{sec:s04}) that investigates the prevalence of some of these atom candidates in large scale, professionally-developed JavaScript software.

Table~\ref{tab:atomsBoth} presents the atom candidates that we explore in all studies, while Table~\ref{tab:atomsRepeated} presents the additional 
ones that only the repeated measures study analyzes.

### 2023-02-21 s01.tex
endereçando os comentários de R1 e R3 após o segundo round de revisões

% The design of the repeated measures study considers two treatments: the presence or absence of atom candidates within the programs.
To identify atom candidates, we start out by selecting every atom candidate from Gopstein's study that can also happen in JavaScript. That excludes atoms related to pointers and preprocessor directives. Furthermore, we conduct an informal search of \revised{JavaScript questions on StackOverflow about constructs that are not typical of other languages, such as array and object spread, solutions to Code Golf~\footnote{https://codegolf.stackexchange.com/} problems, and code style guides\footnote{AirBNB JavaScript Style Guide: https://github.com/airbnb/javascript}\footnote{Google JavaScript Style Guide: https://google.github.io/styleguide/jsguide.html} to identify new atom candidates. We leverage previous work~\cite{castor2018} on the identification of atom candidates to guide us in this informal search.} Overall, the repeated measures study analyzes 24 atoms, 15 from the original study and 9 that are new (see Tables~\ref{tab:atomsBoth} and ~\ref{tab:atomsRepeated}.

@@

{\bf Study instrument.} The experiment is conducted by means of a questionnaire made available as a web application. As part of this effort, we carried out an informal pilot whose main objectives were (i) to spot bugs in the application and in the data collection mechanism; (ii) to gain feedback from respondents about the user experience of the application; and (iii) to formulate an estimate about how long answering the questionnaire would take on average. Based on the feedback of the participants of the pilot, we present only one obfuscated and one clean version of each atom candidate, to reduce to total time of the study and potentially increase participation.% Initially, our plan was to present two of each.

>>>>

% The design of the repeated measures study considers two treatments: the presence or absence of atom candidates within the programs.
To identify atom candidates, we start out by selecting every atom candidate from Gopstein's study that can also happen in JavaScript. That excludes atoms related to pointers and preprocessor directives. Furthermore, we conduct an informal search of \revised{JavaScript questions on StackOverflow about constructs that are not typical of other languages, such as array and object spread, solutions to Code Golf~\footnote{https://codegolf.stackexchange.com/} problems, and code style guides\footnote{AirBNB JavaScript Style Guide: https://github.com/airbnb/javascript}\footnote{Google JavaScript Style Guide: https://google.github.io/styleguide/jsguide.html} to identify new atom candidates. We leverage previous work~\cite{castor2018} on the identification of atom candidates to guide us in this informal search.} Overall, the repeated measures study analyzes 24 atoms, 15 from the original study and 9 that are new (see Tables~\ref{tab:atomsBoth} and ~\ref{tab:atomsRepeated}).

@@

{\bf Study instrument.} The experiment is conducted by means of a questionnaire made available as a web application. As part of this effort, we carried out an informal pilot whose main objectives were (i) to spot bugs in the application and in the data collection mechanism; (ii) to gain feedback from respondents about the user experience of the application; and (iii) to formulate an estimate about how long answering the questionnaire would take on average. Based on the feedback of the participants of the pilot, we present only one obfuscated and one clean version of each atom candidate, to reduce the total time of the study and potentially increase participation.% Initially, our plan was to present two of each.

### 2023-02-21 s02.tex
endereçando os comentários de R1 e R3 após o segundo round de revisões

\subsection{Experimental design}

@@

For each of the 10 selected atom candidates, we write one short program containing the atom. As in the repeated measures study, we call them the obfuscated versions of the programs. We also write corresponding short programs without the atom candidates and call them the clean versions of the programs. Overall, the experiment employed 20 programs, 10 obfuscated and 10 clean. In order to reduce the cognitive effort, each subject was asked to predict the output of 10 programs, five obfuscated and five clean. The order in which they are presented is randomized. By doing this, we seek to minimize the chances of subjects being aware that the current listing they are analyzing contains (or not) atoms of confusion. That is, each subject should indicate what would be the outcomes of the programs, some of them having atoms of confusion (while other programs do not). Since participants are not exposed to obfuscated and clean versions corresponding to the same atom candidate, learning effect is not possible. We measure answer correctness and the total time each participant requires to participate in the experiment.

@@

\subsection{Results}\label{sec:latin}

@@

\subsubsection{Correctness Analysis}
% \castor{We need to be uniform about terminology. Are we going with ``misunderstanding rate analysis''? The problem is that it is not clear to me to which rate this refers.}

{\bf Exploratory Data Analysis.}
As mentioned in Section~\ref{sec:meth:survey}, each participant in this experiment evaluated ten code snippets, from which five were in their obfuscated versions, whilst the other five contained clean versions of the code snippets (i.e., without the atom candidates). As in the repeated measures study, the participants should provide the expected outcomes of the code snippets. Differently from that study, in this one each participant only predicts the outcome of one version of each code snippet, either clean or obfuscated. As discussed, we collect information about \emph{correctness} (whether the participant correctly predicted the program's output) and \emph{time}. 
Table~\ref{tab:difference-correctness} and Figure~\ref{fig:boxplotcorrectness} summarize the results of the correctness experiment.

>>>>

\subsection{Experimental design}\label{sec:latin-design}

@@

For each of the 10 selected atom candidates, we write one short program containing the atom. As in the repeated measures study, we call them the obfuscated versions of the programs. We also write corresponding short programs without the atom candidates and call them the clean versions of the programs. Overall, the experiment employed 20 programs, 10 obfuscated and 10 clean. In order to reduce the cognitive effort, each subject was asked to predict the output of 10 programs, five obfuscated and five clean. The order in which they are presented is randomized. By doing this, we seek to minimize the chances of subjects being aware that the current listing they are analyzing contains (or not) atom candidates. That is, each subject should indicate what would be the outcomes of the programs, some of them having obfuscated code (while other programs do not). Since participants are not exposed to obfuscated and clean versions corresponding to the same atom candidate, learning effect is not possible. We measure answer correctness and the total time each participant requires to participate in the experiment.

@@

\subsection{Results}\label{sec:latin-results}

@@

\subsubsection{Correctness Analysis}\label{sec:latin:correct}
% \castor{We need to be uniform about terminology. Are we going with ``misunderstanding rate analysis''? The problem is that it is not clear to me to which rate this refers.}

{\bf Exploratory Data Analysis.}
As mentioned in Section~\ref{sec:latin-design}, each participant in this experiment evaluated ten code snippets, from which five were in their obfuscated versions, whilst the other five contained clean versions of the code snippets (i.e., without the atom candidates). As in the repeated measures study, the participants should provide the expected outcomes of the code snippets. Differently from that study, in this one each participant only predicts the outcome of one version of each code snippet, either clean or obfuscated. As discussed, we collect information about \emph{correctness} (whether the participant correctly predicted the program's output) and \emph{time}. 
Table~\ref{tab:difference-correctness} and Figure~\ref{fig:boxplotcorrectness} summarize the results of the correctness experiment.

### 2023-02-21 s03.tex
presenting an overview of the studies
review os s04
endereçando os comentários de R1 e R3 após o segundo round de revisões

\section{Interview Study}\label{sec:method:interview}

To complement the two experiments, we perform semi-structured interviews with professional JavaScript developers, aiming to identify their perceptions regarding programs containing atoms of confusion (research question RQ2). We also ask each participant if they know of any other JavaScript-specific construct or idiom they think is likely to make the code hard to understand. % (research question RQ3). 
In this section we detail the protocol we followed to conduct the interviews and to analyze the results.

@@

In this section we present the results of the interviews with the practitioners. We contrast the findings of the experiments of Sections~\ref{sec:repeated} and~\ref{sec:latin} with the opinion of software developers about their preferences regarding the obfuscated or clean versions of the code snippets. For the interview study, we only considered atoms that were analyzed in both the repeated measures (Section~\ref{sec:repeated}) and the latin square (Section~\ref{sec:latin}) studies.

@@

Supporting the results of Sections~\ref{sec:repeated} and~\ref{sec:latin}, Table~\ref{tab:interview-results1}
 shows that for eight out of the
10 scenarios surveyed, the majority of the respondents prefer the version of the
code without the atom candidate.
In no case the \emph{neutral}
ratio was higher than the option for the clean version.
%% An example entry is contained in Appendix~\ref{}. 
%% For a full listing of the code snippets, visit the paper repository. 
Only for the Conditional Operator atom candidate the participants preferred the obfuscated version, instead of the clean one. Figure~\ref{code:ternary} shows the code snippets for the corresponding obfuscated and clean versions. For this atom candidate, some participants who preferred the left-hand side version (obfuscated) still believed that the right-hand side version (clean) was more readable. The following quotes were extracted from the transcripts with
% three interviewees:
two interviewees:

@@

\noindent\begin{minipage}{.45\textwidth}
\begin{lstlisting}[language=JavaScript, caption=\emph{Left-hand side} (using the \emph{Conditional Operator} atom).]
let config = {size: 3, isActive: false};
const_config = config.isActive === true 
             ? config 
             : {size: 10};
console.log(_config.size);
\end{lstlisting}
\end{minipage}\hfill
\begin{minipage}{.45\textwidth}
\begin{lstlisting}[language=JavaScript, caption=\emph{Right-hand side} (without the atom).]
let config = {size: 3, isActive: false}
let _config;
if(config.isActive === true) {
  _config = config;
}
else{
   _config = {size: 10};
}
console.log(_config.size);
\end{lstlisting}
\end{minipage}
\caption{Example of a pair of code snippets used in the interview pertaining to the \emph{Conditional Operator} atom candidate}
\label{code:ternary}
\vspace{-0.2cm}
\end{figure*}

@@

\begin{mq}
  \emph{``This one is interesting, because I have written code that looks like the left-hand side [obfuscated version], and my colleagues complained that it was difficult to understand. Nowadays I prefer to write code using the version on the right-hand side [clean version]''}.
\end{mq}
\noindent
None of the two studies found a significant difference in correctness between clean and obfuscated versions of snippets related to this atom candidate. However, in the latin square study the participants analyzing clean versions of the snippets required significantly less time.

>>>>

\section{Interview Study}\label{sec:s03}

To complement the two experiments, we perform semi-structured interviews with professional JavaScript developers, aiming to identify their perceptions regarding programs containing atoms of confusion (research question RQ3). We also ask each participant if they know of any other JavaScript-specific construct or idiom they think is likely to make the code hard to understand. % (research question RQ3). 
In this section we detail the protocol we followed to conduct the interviews and to analyze the results.

@@

In this section we present the results of the interviews with the practitioners. We contrast the findings of the experiments of Sections~\ref{sec:s01} and~\ref{sec:s02} with the opinion of software developers about their preferences regarding the obfuscated or clean versions of the code snippets. For the interview study, we only considered atoms that were analyzed in both the repeated measures (Section~\ref{sec:s01}) and the latin square (Section~\ref{sec:s02}) studies.

@@

Supporting the results of Sections~\ref{sec:repeated} and~\ref{sec:latin-results}, Table~\ref{tab:interview-results1}
 shows that for eight out of the
10 scenarios surveyed, the majority of the respondents prefer the version of the
code without the atom candidate.
In no case the \emph{neutral}
ratio was higher than the option for the clean version.
%% An example entry is contained in Appendix~\ref{}. 
%% For a full listing of the code snippets, visit the paper repository. 
Only for the Conditional Operator atom candidate the participants preferred the obfuscated version, instead of the clean one. Figure~\ref{code:ternary} shows the code snippets for the corresponding obfuscated and clean versions. For this atom candidate, some participants who preferred the left-hand side version (obfuscated) still believed that the right-hand side version (clean) was more readable. The following quotes were extracted from the transcripts with
% three interviewees:
two interviewees:

@@

%\noindent\begin{minipage}{.45\textwidth}
\begin{lstlisting}[language=JavaScript, caption=\emph{Left-hand side} (using the \emph{Conditional Operator} atom).]
let config = {size: 3, isActive: false};
const_config = config.isActive === true 
             ? config 
             : {size: 10};
console.log(_config.size);
\end{lstlisting}
%\end{minipage}\hfill
%\begin{minipage}{.45\textwidth}
\begin{lstlisting}[language=JavaScript, caption=\emph{Right-hand side} (without the atom).]
let config = {size: 3, isActive: false}
let _config;
if(config.isActive === true) {
  _config = config;
}
else{
   _config = {size: 10};
}
console.log(_config.size);
\end{lstlisting}
%\end{minipage}
\caption{Example of a pair of code snippets used in the interview pertaining to the \emph{Conditional Operator} atom candidate}
\label{code:ternary}
\vspace{-0.2cm}
\end{figure*}

@@

\begin{mq}
  \emph{``This one is interesting, because I have written code that looks like the left-hand side [obfuscated version], and my colleagues complained that it was difficult to understand. Nowadays I prefer to write code using the version on the right-hand side [clean version]''}.
\end{mq}
\noindent
None of the two experiments we detailed in Sections~\ref{sec:s01} and~\ref{sec:s02} found a significant difference in correctness between clean and obfuscated versions of snippets related to the Logic as Control Flow atom candidate. However, in the latin square study the participants analyzing clean versions of the snippets required significantly less time to predict a correct answer.
